Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-11-30 analysis/paper/paper.qmd
edited up to Principal Component Analysis

All data preparation, analyses and visualization were performed in the R environment [@RCoreTeam]. Our R code and data, including the original artefact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].

@@

@fig-plot-cv-over-time shows that the directional trend in CV values of artefact attributes is complicated. Half of the attributes are below our threshold value of `r threshold_value`, and half are above. Only body length crosses the threshold value, changing from <`r threshold_value` to >`r threshold_value` over time, indicating a shift from indirect bias to guided variation. For those attributes where the CV is <`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes. For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over times, suggesting increased influence of guided variation for these attributes.

@@

One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-over-time). These four assemblages are more likely to represent a consistent way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making. The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS). SYG1_2 and SYG6_2 are from Phase 2 (shown on the left side of the figure) and SYG1_3 and YS_3 are from Phase 3 (shown on the right side of the figure). Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly below `r threshold_value`, which represent transmissions dominated by indirect bias. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-over-time. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of indirect bias, rather than substantial changes in the type of bias dominating cultural transmission.

@@

We examined the CV values per site to address our research question about the cultural transmission over space. We excluded assemblages that contain a single point. @fig-plot-cv-by-site shows that Sachang site has the highest CV values with the widest confidence level. Sachang is part of Phase 1 and its CV values explain why the CV of all time is higher than Phase 2. Two assemblages from the Suyanggae site have relatively stable CV for attributes, which indicate highly standardized shapes of stemmed points. CV values of body length is the lowest while stem width the highest among all assemblages. Overall, there are no clear patterns between the sites.

```{r}
#| label: fig-plot-cv-by-site
#| fig-cap: CV values per site that has more than one stemmed point. The points represent CV values for each attribute. The vertical lines indicate the range of confidence intervals.
#| fig-width:   8.5
#| fig-height:  4.5

cv_by_site_df  <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage, 
                -lat_dd, 
                -long_dd, 
                -sitename, 
                -SPstage.Raw_material) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  dplyr::filter( n > 1) %>%
  dplyr::select(-n) %>%
  pivot_longer(-full_sitename,
               names_to = "variable",
               values_to = "value") %>%
  group_by(full_sitename, variable) %>%
  dplyr::summarise(cv_by_site =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

@@

cv_plot_site_label %>%
  ggplot(aes(x = variable,
             y = cv_by_site)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma),
                width = 0.2) +
  xlab("Attribute") +
  ylab("Coefficient of Variation on Attributes") +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  facet_wrap( ~ label, scales = "free_y") +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 6.5))
```

@@

We examined the relationship between CV values and raw materials with a premise that the shape of stemmed points are highly dependent on raw materials. We excluded raw materials that were used for single point. @fig-plot-cv-by-raw-material shows that quartzite, the raw materials for Sachang stemmed points, has the highest CV value with a wide range of conference intervals. Quartz has high variations of CV among the attributes. On the other hand, shale has the lower CV, which is very stable throughout the attributes.

@@

# CV by raw materials
### Make a table of CVs for all variable grouped by raw materials
cv_by_raw_materials <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage, -lat_dd, -long_dd, -sitename, -full_sitename) %>%
  group_by(SPstage.Raw_material) %>%
  add_tally() %>%
  dplyr::filter( n > 1) %>%
  dplyr::select(-n) %>%
  pivot_longer(-SPstage.Raw_material,
               names_to = "variable",
               values_to = "value") %>%
  group_by(SPstage.Raw_material, variable) %>%
  dplyr::summarise(cv_by_materials =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

>>>>

All data preparation, analyses and visualization were computed in the R environment [@RCoreTeam]. Our R code and data, including the original artefact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].

@@

@fig-plot-cv-over-time shows that the directional trend in CV values of artefact attributes is complicated. Half of the attributes are below our threshold value of `r threshold_value`, and half are above. Only body length crosses the threshold value, changing from <`r threshold_value` to >`r threshold_value` over time, indicating a shift from indirect bias to guided variation. For those attributes where the CV is <`r threshold_value`, the trend is decreasing CV values from Phase 2 to Phase 3 for maximum width and tang length, indicating increasing influence of indirect bias for these attributes. For stem length, stem width and tang length, CV values remain above `r threshold_value`, and show a slight trend to increase over time, suggesting increased influence of guided variation for these attributes.

@@

One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to a local tradition of artefact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-over-time). These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artefact making. The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS). SYG1_2 and SYG6_2 are from Phase 2 (shown on the left side of the figure) and SYG1_3 and YS_3 are from Phase 3 (shown on the right side of the figure). Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly below `r threshold_value`, which represent transmissions dominated by indirect bias. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-over-time. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of indirect bias, rather than substantial changes in the type of bias dominating cultural transmission.

@@

We examined the CV values per site to address our research question about the cultural transmission over space. We excluded assemblages that contain less than five points. @fig-plot-cv-by-site shows higher CV values for stem length and stem width, indicating high variability at the stem of the points from these sites. Overall size variability is similar between these three assemblages, suggesting a consistent approach to constraining shape variation over space, rather than site-specific variances.

```{r}
#| label: fig-plot-cv-by-site
#| fig-cap: CV values per site that has more than five stemmed points. The points represent CV values for each attribute. The vertical lines are confidence intervals.
#| fig-width:   8.5
#| fig-height:  4.5

cv_by_site_df  <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage, 
                -lat_dd, 
                -long_dd, 
                -sitename, 
                -SPstage.Raw_material) %>%
  group_by(full_sitename) %>%
  add_tally() %>%
  dplyr::filter( n > 5) %>%
  dplyr::select(-n) %>%
  pivot_longer(-full_sitename,
               names_to = "variable",
               values_to = "value") %>%
  group_by(full_sitename, variable) %>%
  dplyr::summarise(cv_by_site =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))

@@

cv_plot_site_label %>%
  ggplot(aes(x = variable,
             y = cv_by_site)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma),
                width = 0.2) +
  xlab("Attribute") +
  ylab("Coefficient of Variation on Attributes") +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  facet_wrap( ~ label, scales = "free_y") +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0,100)) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 6.5))
```

@@

We examined the relationship between CV values and raw materials to test the hypothesis that the shape of stemmed points was highly dependent on raw material. We excluded raw materials that were used for less than three points. @fig-plot-cv-by-raw-material shows that quartzite, the raw materials for stemmed points at Sachang, has the highest CV value with a wide range of conference intervals. Other raw materials are generally stable in variation throughout the attributes.

@@

# CV by raw materials
### Make a table of CVs for all variable grouped by raw materials
cv_by_raw_materials <-
  df_full_sitename %>%
  dplyr::select(-SPstage.Stage, -lat_dd, -long_dd, -sitename, -full_sitename) %>%
  group_by(SPstage.Raw_material) %>%
  add_tally() %>%
  dplyr::filter( n >= 3) %>%
  dplyr::select(-n) %>%
  pivot_longer(-SPstage.Raw_material,
               names_to = "variable",
               values_to = "value") %>%
  group_by(SPstage.Raw_material, variable) %>%
  dplyr::summarise(cv_by_materials =   sharma_cv(value),
            cv_low_sharma =  sharma_int_low(value),
            cv_high_sharma = sharma_int_high(value))
